Multiresolution Neural Field Estimation

### 2011-02-28 Multiresolution_framework/ltx/manuscript.tex
resolved the conflict
updated the plots

The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used to in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Gibsona2005,Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For a multivariate linear dynamic system, the bound used is a quadratic of the form
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
which has a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ block of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\mike{you need to define and sort out the $[]_{ij}$ notation!}The $\left(i,j\right)^{th}$ element of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix.\cut{The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.} The matrices $\Xi_0$ and $\Xi_1$ are calculated using the Rauch Tung Streibel \cite{Rauch}(RTS) smoother outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_t,\mathbf{x}_{t-1})$ 
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}

Hence, the algorithm has two distinct steps: the E-step which computes $\Xi_0$ and $\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step which updates the paramter estimates by computing the (analytic) maximum of $Q(\theta,\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge; multiple restarts are required to avoid convergence on a local maxima.

@@

The spatial frequency of the observed field is shown in \figurename{\ref{fig:KernelAndFreqResponse}} (b), showing a spatially band-limited signal. This suggests the use of wavelets up to level $j=3$ (with the bandwidth $\approx[4.8,8]$ cycles/mm) can fully represent the field. Note that wavelets due to their bandpass filter characteristics extract successively lower and lower frequency components.

Field reconstruction \dean{using the results of the EM algorithm} at different levels of approximations is shown in \figurename{\ref{fig:FieldEstimates}}. \dean{Needs more info.} The figure confirms that $j=3$ is adequate to capture the dynamic of the underlying field \dean{since the error converges?}. The root mean square error (RMSE) of the field estimates is shown in \figurename{\ref{fig:RMSE}}.

The histogram in \figurename{\ref{fig:ParametersHistogram}} demonstrates the performance of the parameter estimation. \dean{The results are good because the RMSE between the true and estimated kernels is $1.83$ (units?)... There is a bias of 7.4 \% in $\theta_1$ because..., which is not big deal.}
\begin{figure}[!h]
 \centering
 \includegraphics[scale=1]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations generated by the kernel in (a). The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 %\includegraphics[width=0.5\textwidth]{./Graph/RMSE.pdf}
\includegraphics[scale=1]{./Graph/RMSE.pdf}
 \caption{RMSE of the estimated field at different spatial resolution.}
 \label{fig:RMSE}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Field.pdf}
 \caption{Spatial field at one time instant for different spatial resolutions, estimated and true fields are shown by dashed and solid lines respectively; (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{Histograms of the parameter estimates over 100 realizations. In each of the subplots
the solid lines show the actual parameter values and the dotted lines show the means of the estimated parameters. (a) The histogram of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. (b) The histogram of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$.}
 \label{fig:ParametersHistogram}
 \end{figure}

>>>>

The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used to in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Gibsona2005,Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For a multivariate linear dynamic system, the bound used is a quadratic of the form
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
which has a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ block of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\mike{you need to define and sort out the $[]_{ij}$ notation!}
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix.\cut{The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.} The matrices $\Xi_0$ and $\Xi_1$ are calculated using the Rauch Tung Streibel (RTS)\cite{RAUCH1965} smoother outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_t,\mathbf{x}_{t-1})$ 
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}

Hence, the algorithm has two distinct steps: the E-step which computes $\Xi_0$ and $\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step which updates the paramter estimates by computing the (analytic) maximum of $Q(\theta,\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge; \parham{ we dont have local maxima here?} multiple restarts are required to avoid convergence on a local maxima.

@@

The spatial frequency of the observed field is shown in \figurename{\ref{fig:KernelAndFreqResponse}} (b), showing a spatially band-limited signal. This suggests the use of wavelets up to level $j=2$ (with the bandwidth $\approx[2.5,4]$ cycles/mm) can fully represent the field. Note that wavelets due to their bandpass filter characteristics extract successively higher and higher frequency components by increasing $j$.

Field reconstruction \dean{using the results of the EM algorithm} at different levels of approximations is shown in \figurename{\ref{fig:FieldEstimates}}. \dean{Needs more info.} The figure confirms that $j=2$ is adequate to capture the dynamic of the underlying field \dean{since the error converges?}. The root mean square error (RMSE) of the field estimates is shown in \figurename{\ref{fig:ParametersHistogram}} (a).

The histogram in \figurename{\ref{fig:ParametersHistogram}} demonstrates the performance of the parameter estimation. \dean{The results are good because the RMSE between the true and estimated kernels is $1.83$ (units?)... There is a bias of 7.4 \% in $\theta_1$ because..., which is not big deal.}
\begin{figure}[!h]
 \centering
 \includegraphics[scale=1]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations generated by the kernel in (a). The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Field.pdf}
 \caption{Spatial field at one time instant for different spatial resolutions, estimated and true fields are shown by dashed and solid lines respectively; (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{(a) RMSE of the estimated field at different spatial resolution, $j$. (b) The histogram of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. (c) The histogram of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$. Histograms are obtained over 100 realizations. In each of the histograms
the solid lines show the actual parameter values and the dotted lines show the means of the estimated parameters.}
 \label{fig:ParametersHistogram}
 \end{figure}
